---
title: "trnF data analysis using DADA2 part-2"
author: "Pedro Rodrigues"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    fig_caption: yes

fontsize: 14pt
---

# Part 2, adding results from blast search


Prepare environment:
```{r init, warning=FALSE, message=FALSE}

setwd("/scratch/pd88715/ref_seq_plastids")
set.seed("1805")

library(dada2);packageVersion("dada2")
library(Biostrings); packageVersion("Biostrings")
library(ShortRead); packageVersion("ShortRead")
library(ggplot2); packageVersion("ggplot2")
library(reshape2); packageVersion("reshape2")
library(gridExtra); packageVersion("gridExtra")
library(phyloseq); packageVersion("phyloseq")
library(dplyr); packageVersion("dplyr")
```

```{r more preparation, warning=FALSE, message=FALSE}
pathPBB1 <- "../BurrowerBugs/PBB1"
pathPBB2 <- "../BurrowerBugs/PBB2"
pathPBB3 <- "../BurrowerBugs/PBB3" # CHANGE ME to location of the First Replicate fastq files
path.out <- "Figures/"
path.rds <- "RDS/"
theme_set(theme_bw())
```

Load Data

```{r phyloseq-load-improved-data}
p_data.cleanOTU <- readRDS("RDS/p_data.cleanOTU.rds")
p_data.cleanSampleData <- readRDS("RDS/p_data.cleanSampleData.rds")
p_data.cleanOTUTaxaTable <- readRDS("RDS/p_data.cleanOTUTaxaTable.rds")
p_data.cleanOTURefSeq <- readRDS("RDS/p_data.cleanOTURefSeq.rds")
```

Make Phyloseq object
```{r phyloseq-make-object}
p_data.clean <- phyloseq(otu_table(p_data.cleanOTU, taxa_are_rows = FALSE), 
                   sample_data(p_data.cleanSampleData),
                   tax_table(p_data.cleanOTUTaxaTable),
                   refseq(p_data.cleanOTURefSeq))
```


>Blast was executed using script under the file name blast_ASV_pbb123.sh
>PBB123_asv_03292023.csv (taxonomy table) was edited and every NA was changed if taxonomic information was available (a blast hit result)
>NOTE: taxonomical assignments ranged from about 95% to 100% and some NAs remained with no classification via Blast
#### Add new taxonomical information to Phyloseq object

```{r add-new-taxon-table}
new_tax_table <- read.csv("PBB123_asv_03292023_withBlastids.csv", row.names = 1)
new_tax <- tax_table(as.matrix(new_tax_table))
tax_table(p_data.clean) <- new_tax
```

## Exploratory analysis

List all instances where Genus == NA
```{r list-genus-NA}
#tax_table(p_data.clean) %>% as("matrix") %>% as_tibble 
p_data.clean %>%
  subset_taxa(is.na(Genus)) %>%
 tax_table()
```

All but two instance of Genus=NA are ASV where Kingdom is also NA. As a conservative approach, I will exclude everything that is not identified as Viridiplantae at the Kingdom level


=-=-=-=-=-=-=- Legacy stuff - ignore - =-=-=-=-=-=-=-=-=-=-=-=-
```{r fix-names-and-list-duplicate-samples}
# ### NOTE: duplicates were found to be actual different samples; skip this
#sample_data(p_data.clean)$Indiv_ID <- sub("_$","",sample_data(p_data.clean)$Indiv_ID)
#find_duplicates <- sample_data(p_data.clean) %>% as("matrix") %>% as_tibble %>% select(c(Indiv_ID, Duplicated)) %>% arrange(Indiv_ID)
```

There is one instance where one sample lost its duplicate; in this case, I will change the column "duplicated" to "no" for sample MSC_33_2
```{r fix-sample-MSC_33_2}
# ### NOTE: duplicates were found to be actual different samples; skip this
#find row at where sample is located 
#order_original <- sample_data(p_data.clean) %>% as("matrix") %>% as_tibble() %>% select(c(Indiv_ID, Duplicated)) #row 9
# change value in column "Duplicated" to "no"
#sample_data(p_data.clean)$Duplicated[9] <- "no"
#sample_data(p_data.clean) %>% as("matrix") %>% as_tibble %>% select(c(Indiv_ID, Duplicated)) %>% head(., n=10)
```
=-=-=-=-=-=-=-=-=-==-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=


Remove singletons if needed and taxa where Order = NA
Remove instances where algae was identified
Remove all duplicated samples #No duplicates; skip
Remove samples where count is zero
```{r prepare-for-downstream-diversity-analysis}
tail(taxa_sums(p_data.clean) < 2)
p_data.clean <- prune_taxa(taxa_sums(p_data.clean) > 1, p_data.clean) %>% subset_taxa(., Kingdom == "Viridiplantae")
#p_data.clean <- subset_samples(p_data.clean, Duplicated=="no")
p_data.clean <- subset_samples(p_data.clean, sample_sums(p_data.clean) > 0)
```

Sanity check - is there any NA left for Genus?
```{r list-genus-NA-2}
#tax_table(p_data.clean) %>% as("matrix") %>% as_tibble 
p_data.clean %>%
  subset_taxa(is.na(Genus)) %>%
 tax_table()
```

only one ASV is NA (no hits via blast either)

```{r save-progress}
saveRDS(otu_table(p_data.clean), "RDS/p_data.cleanOTU_cleaningcompleted.rds")
saveRDS(sample_data(p_data.clean), "RDS/p_data.cleanSampleData_cleaningcompleted.rds")
saveRDS(tax_table(p_data.clean), "RDS/p_data.cleanOTUTaxaTable_cleaningcompleted.rds")
saveRDS(refseq(p_data.clean), "RDS/p_data.cleanOTURefSeq_cleaningcompleted.rds")
```